
---
title: "Temporal Aggregation of (Growth) Timeseries"
author: "Rob van Harrevelt and Anita van der Roest"
date: '`r Sys.Date()`'
output:
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
linkcolor: blue    
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{Temporal Aggregation of (Growth) Timeseries}
-->

\newcommand{\longerpage}[1][1]{\enlargethispage{#1\baselineskip}}
\newcommand{\shorterpage}[1][1]{\enlargethispage{-#1\baselineskip}}
\newcommand{\needlines}[1]{\Needspace{#1\baselineskip}}


```{r echo = FALSE, message = FALSE}
library(regts)
library(Hmisc)
library(readxl)
data <- matrix(c(3,2,4,3,5,2,5,3), ncol = 2)
data1 <- matrix(c(1,3,2,6,4,8,3,0,9), ncol = 3)

# input/output chunks are separated
# R inputlines start with >, outputlines with nothing
knitr::opts_chunk$set(collapse = FALSE, comment = "", prompt = TRUE)

# set options for page width = 100
options(width = 100)

```

# Introduction
This vignette discusses temporal aggregation of timeseries, i.e. the conversion 
of a timeseries to a lower frequency. 
The vignette is an extension of the vignette [_"Introduction to regts"_](regts.pdf), but can be read independently.

Section 2 describes the standard aggregation function `aggregate`. Section 3 discusses aggregation for growth timeseries with several methods, which are described in detail in section 4.


# Function aggregate
The standard function `aggregate` can be used to convert a timeseries to a lower 
frequency, for example a quarterly to an annual timeseries, or a monthly to a quarterly timeseries.

An example with a monthly timeseries:
```{r}
regm <- regts(1:24, start = "2016M1")
regm
aggregate(regm, FUN = mean, nfrequency = 4)
```
The observations in the timeseries are split into blocks for each quarter and `FUN` is applied to the three observations in that quarter. The result is a quarterly timeseries holding the aggregated values.

Another example with a quarterly timeseries:
```{r}
regq <- regts(1:10, start = "2016q1")
regq
aggregate(regq, FUN = sum)
```
The quarters in 2016 add up to 10, the quarters in 2017 to 26. Note that the two 
quarters in 2018 are ignored. The `aggregate` function skips the incomplete 
years at the end.

Incomplete years at the beginning are also skipped. For example, the next timeseries starts in the second quarter:

```{r}
regq2 <- regts(2:10, start = "2016q2")
regq2
aggregate(regq2, FUN = sum)
```

The quarters in 2017 add up to 26 as before. But now the information for 2016 
is also not complete, and therefore ignored. 

Thus for `regts` objects the `aggregate` function skips all incomplete years at the beginning and end of the timeseries. However for standard timeseries (`ts`) the implementation of `aggregate` only skips incomplete years at the end.
The previous example but now with a `ts` timeseries:

```{r}
tq2 <- ts(2:10, start = c(2016,2), frequency = 4)
tq2
aggregate(tq2, FUN = sum)
```

\longerpage[1]
The result is an unconventional timeseries with frequency `year` but a shifted period (the start is 2016.25). The result for 2016.25 is the sum of the observations in period `2016Q2/2017Q1`. 

In all other cases the `aggregate` function works exactly the same for `regts` and `ts` objects.

\newpage
# Aggregation of growth timeseries 
As is shown in the next section the function `aggregate` does not yield correct 
results for timeseries with absolute, relative or percentage changes. 
For growth timeseries the function `aggregate_gr` can be used. It accepts both 
`regts` and `ts` timeseries. The function implements several methods 
(`rel`, `pct`, `difmean` and `difsum`) for different types of input timeseries.

The `rel` and `pct` methods assume that the input timeseries is a one-period 
relative or percentage change and calculate the exact low frequency relative or 
percentage change. Methods `difmean` and `difsum` assume that the input timeseries 
is a first difference. 
Some examples for change timeseries, the results are (by default) annual timeseries:
An example for a percentage change series:
```{r eval = FALSE}
xgrowth <- growth(regts(1:8, start = "2019q1"))
aggregate_gr(xgrowth, method = "rel")
```

Examples for difference series:
```{r}
xdif <- diff(regts(1:8, start = "2019q1"))
xdif
aggregate_gr(xdif, method = "difmean")
aggregate_gr(xdif, method = "difsum")  
```
Note that the results differ factor 4.

The methods are described in detail in the next section.

# Aggregation methods for growth timeseries
This section describes the methods implemented in function `aggregate_gr`, 
including the derivation of the corresponding mathematical formulas. 

Suppose we are temporally aggregating a timeseries $x$ with $n$ subperiods to a 
timeseries $X$ with a lower frequency. In other words we want to convert a 
timeseries x with a high frequency (monthly, quarterly) to a new timeseries with 
a lower frequency (quarterly, annual).

Let $x_{t,i}$   stand for the value of $x$ in subperiod $i$ of main period $t$.
The index $i$ takes on values in the range $1 .. n$.
We always interpret $x_{t,0}$ to mean $x_{t-1,n}$.
The time index $t$ refers to the periods in the time domain of $X$.
Many conversion methods only need the observations subperiod $i$ of main period $t$.
For example the \texttt{mean} method calculates $X_t$ as follows
\begin{displaymath}
X_t = \sum_{i=1}^{n} x_{t,i}/n
\end{displaymath}

However there are transformations which are not so straightforward. A case in point 
is when $x$ is a high frequency first difference and must be converted to a low 
frequency first difference. A simple averaging of $n$ high frequency observations 
will not give the correct answer. For example, for a quarterly growth series the 
yearly aggregates also depend on the quarterly observations of the previous year. 
Therefore we cannot use function `aggregate`, function `aggregate_gr` must be 
applied. In the next subsections a more detailed description of the available 
frequency conversion methods for this function is given.

## `difsum` method 
Define $x$ to be the first difference  of $z$. Thus
\begin{displaymath}
 x_{t,i} = z_{t,i} - z_{t,i-1}
\end{displaymath}
We can also define $z_{t,i}$ in terms of $x$ as follows
\begin{eqnarray*}
z_{t,i} & = & z_{t,i-1} + x_{t,i}\\
        & = & z_{t-1,n} + \sum_{j=1}^{i} x_{t,j}
\end{eqnarray*}
Now define $Z_t$ as the level timeseries in time domain $t$ corresponding
to $z$
\begin{eqnarray*}
Z_t &=& \sum_{i=1}^{n} z_{t,i}\\
    &=& \sum_{i=1}^{n} z_{t-1,n} + \sum_{i=1}^{n} \sum_{j=1}^{i} x_{t,j}\\
    &=& \sum_{i=1}^{n} z_{t-2,n} + \sum_{i=1}^{n} \sum_{j=1}^{n} x_{t-1,j} + \sum_{i=1}^{n} \sum_{j=1}^{i} x_{t,j}
\end{eqnarray*}
Then $X$ in any period $t$ can be calculated from
\begin{displaymath}
X_t = Z_t - Z_{t-1}
\end{displaymath}
Since $Z_{t-1}$ can be written as
\begin{displaymath}
Z_{t-1} = \sum_{i=1}^{n} z_{t-2,n} + \sum_{i=1}^{n} \sum_{j=1}^{i} x_{t-1,j}\\
\end{displaymath}
it is easy to show that $X_t$ does not depend on $z_{t-2,n}$.
In any practical algorithm $z_{t-2,n}$ can be set to 0.

A compact equation for $X_t$ can be derived by substituting the equations for
$Z_t$ and $Z_{t-1}$ into the definition of $X_t$
\begin{displaymath}
X_t = \sum_{i=1}^n
        \left(
          \sum_{j=i+1}^n x_{t-1,j} + \sum_{j=1}^i x_{t,j}
        \right) 
\end{displaymath}
This expression can be futher simplified by changing the order of summation.
For the first summation, we can write
\begin{displaymath}
 \sum_{i=1}^n  \sum_{j=i+1}^n x_{t-1,j} = \sum_{j=2}^n \sum_{i=1}^{j-1} x_{t-1,j}
                    = \sum_{j=2}^n  (j-1) x_{t-1,j}
\end{displaymath}
Similarly,
\begin{displaymath}
 \sum_{i=1}^n  \sum_{j=1}^i x_{t,j} = \sum_{j=1}^n  \sum_{i=j}^{n} x_{t,j} =
                             \sum_{j=1}^n  (n - j + 1) x_{t,j}
\end{displaymath}
The final equation is given by
\begin{displaymath}
X_t =  \left(  \sum_{j=2}^n  (j-1) x_{t-1,j} +  \sum_{j=1}^n  (n - j + 1) x_{t,j} \right) 
\end{displaymath}


## `difmean` method 
The `difmean` method is similar to the `difsum` method described in the previous 
subsection. The only difference is that  the aggregate $Z_t$ is calculated with
an average instead of a sum:
\begin{eqnarray*}
Z_t &=& \sum_{i=1}^{n} z_{t,i}/n\\
\end{eqnarray*}
Further derivations are analogous.
The result for `difmean` is the result for `difsum` divided by $n$:
\begin{displaymath}
X_t =  \left(  \sum_{j=2}^n  (j-1) x_{t-1,j} +  \sum_{j=1}^n  (n - j + 1) x_{t,j} \right) / n
\end{displaymath}


## `rel` method 
Define $x$ to be the relative change in $z$. Thus, assuming that $z >= 0$,
\begin{displaymath}
 x_{t,i} = (z_{t,i} - z_{t,i-1}) / z_{t,i-1}
\end{displaymath}
We can also define $z_{t,i}$ in terms of $x$ as follows
\begin{eqnarray*}
z_{t,i} & = & z_{t,i-1}(1 + x_{t,i})\\
        & = & z_{t-1,n}  \prod_{j=1}^{i}(1+ x_{t,j})\\
        & = & z_{t-2,n} \prod_{j=1}^{n}(1+ x_{t-1,j}) \prod_{j=1}^{i}(1+ x_{t,j})
\end{eqnarray*}
Now define $Z_t$ as the level timeseries in time domain $t$ corresponding
to $z$
\begin{eqnarray*}
Z_t &=& \sum_{i=1}^{n} z_{t,i}\\
    &=&  z_{t-1,n} \sum_{i=1}^{n} \prod_{j=1}^{i}(1+ x_{t,j})\\
    &=&  z_{t-2,n} \prod_{j=1}^{n}(1+ x_{t-1,j}) \sum_{i=1}^{n} \prod_{j=1}^{i}(1+ x_{t,j})
\end{eqnarray*}
Then $X$ in any period $t$ can be calculated from
\begin{displaymath}
X_t = Z_t / Z_{t-1} - 1
\end{displaymath}
Using the expression for $Z_{t-1}$
\begin{displaymath}
Z_{t-1} = z_{t-2,n}  \sum_{i=1}^{n} \prod_{j=1}^{i}(1+ x_{t-1,j})
\end{displaymath}
we obtain
\begin{displaymath}
X_t = \frac{\prod_{j=2}^{n}(1+ x_{t-1,j}) \sum_{i=1}^{n} \prod_{j=1}^{i}(1+ x_{t,j})}
{1 + \sum_{i=2}^{n} \prod_{j=2}^{i}(1+ x_{t-1,j})} - 1
\end{displaymath}


## `pct` method 
Define $x$ to be the percentage change in $z$. Thus,
as before assuming that $z >= 0$,
\begin{displaymath}
 x_{t,i} = 100 (z_{t,i} - z_{t,i-1}) / z_{t,i-1}
\end{displaymath}

The derivations are analogous to the `rel` method described in the previous subsection. 

The result is
\begin{displaymath}
X_t = 100 \frac{\prod_{j=2}^{n}(1+ x_{t-1,j}/100) \sum_{i=1}^{n} \prod_{j=1}^{i}(1+ x_{t,j}/100)}
{1 + \sum_{i=2}^{n} \prod_{j=2}^{i}(1+ x_{t-1,j}/100)} - 1
\end{displaymath}

<!--
-->













